System Architecture and Design: PyroVelocity

### 2025-05-12 nbs/development/architecture.qmd
docs(architecture): add diagrams, module descs, and roadmap
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Architecture
format:
  html:
    toc: true
---

## Overview

PyroVelocity has a modular architecture that separates concerns into distinct components. This document describes the architecture of the PyroVelocity codebase, focusing on the modular implementation in `src/pyrovelocity/models/modular`.

## Protocol-First Design

PyroVelocity follows a protocol-first architecture with all contracts defined through Protocol interfaces. Components implement these interfaces directly without inheritance, promoting composition over inheritance.

## Core Components

### Model Components

@@

- **Dynamics Models**: Implement the mathematical models of RNA dynamics
  - `StandardDynamicsModel`: The standard implementation of RNA dynamics
  - `LegacyDynamicsModel`: Compatibility implementation matching the legacy model

- **Prior Models**: Define the prior distributions for model parameters
  - `LogNormalPriorModel`: Uses log-normal distributions for parameters

- **Likelihood Models**: Define the observation models
  - `PoissonLikelihoodModel`: Uses Poisson distributions for RNA counts
  - `LegacyLikelihoodModel`: Compatibility implementation matching the legacy model

- **Observation Models**: Handle the interface between data and model
  - `StandardObservationModel`: The standard implementation for observations

- **Guide Factories**: Create guide functions for variational inference
  - `AutoGuideFactory`: Creates guides for the modular implementation
  - `LegacyAutoGuideFactory`: Compatibility implementation matching the legacy guide

### Utility Modules

The architecture includes several utility modules:

- `core_utils.py`: Core utility functions used across components
- `context_utils.py`: Utilities for managing Pyro contexts
- `pyro_utils.py`: Utilities specific to Pyro functionality

## Integration with AnnData

The modular implementation integrates directly with AnnData through methods in the `PyroVelocityModel` class, eliminating the need for adapter classes.

## Validation Framework

The architecture includes a validation framework to ensure that the modular implementation can replicate the behavior of the legacy implementation.

## Future Directions

The architecture is designed to be extensible, allowing for the addition of new components and features while maintaining backward compatibility with existing workflows.

>>>>

---
title: "System Architecture and Design: PyroVelocity"
format:
  html:
    mermaid-format: js
    mermaid:
      theme: dark
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

# Introduction {#sec-introduction}

## Purpose {#sec-purpose}

This document describes the architecture and design of PyroVelocity's modular implementation. It serves as both a reference for ongoing development and as a resource for new developers and users who may be interested in extending the functionality of existing models or creating and comparing new ones. As such, this document will evolve alongside the implementation.

## Scope {#sec-scope}

This document covers:

- High-level system architecture
- Key components and their interactions
- Data flow and processing pipelines
- Major design decisions and their rationales
- Technical constraints and quality attributes

The focus is currently on the modular model implementation in `src/pyrovelocity/models/modular`. Additional sections will be added as their architecture stabilizes.

## Definitions {#sec-definitions}

| Term | Definition |
|------|------------|
| RNA Velocity | The time derivative of gene expression, representing the rate of change of mRNA abundance |
| AnnData | A data structure for annotated data matrices, commonly used in single-cell genomics |
| Protocol | A Python typing construct that defines an interface without implementation |
| Pyro | A probabilistic programming framework built on PyTorch |
| SVI | Stochastic Variational Inference, a method for approximate Bayesian inference |

# System Overview {#sec-system-overview}

## Vision {#sec-vision}

PyroVelocity aims to provide a robust, flexible, and extensible framework for probabilistic RNA velocity analysis. The modular architecture enables researchers to customize and extend the model while maintaining a consistent interface.

## Architectural Style {#sec-architectural-style}

PyroVelocity follows a Protocol-First architecture, chosen to maximize flexibility and composability. This style enables:

1. **Composition over inheritance**: Components implement Protocol interfaces directly without inheritance hierarchies
2. **Loose coupling**: Components interact through well-defined interfaces
3. **Testability**: Components can be tested in isolation
4. **Extensibility**: New components can be added without modifying existing code

## Context Diagram {#sec-context-diagram}

```{mermaid}
graph TB
    User["Data Scientist / Researcher"] -->|uses| Library["PyroVelocity"]
    Library -->|depends on| PyTorch["PyTorch"]
    Library -->|depends on| Pyro["Pyro"]
    Library -->|processes| AnnData["AnnData Objects"]
    Library -->|produces| Velocity["RNA Velocity"]
    Library -->|integrates with| Scanpy["Scanpy"]
    Library -->|integrates with| Scvelo["scVelo"]

    class User user;
    class Library library;
    class PyTorch,Pyro,Scanpy,Scvelo external;
    class AnnData,Velocity data;
```

## Key Principles {#sec-key-principles}

The following principles guide architectural decisions for PyroVelocity:

1. **Protocol-First Design**: All contracts are defined through Protocol interfaces, enabling structural typing and composition over inheritance.
2. **Separation of Concerns**: Each component has a single responsibility, making the system more maintainable and testable.
3. **Immutable State**: The system uses immutable data structures where possible to reduce side effects and improve reasoning about code.
4. **Type Safety**: The system uses static typing and runtime type checking to catch errors early and provide better development support.

# Structural View {#sec-structural-view}

## Component Architecture {#sec-component-architecture}

The modular implementation is composed of the following major components:

```{mermaid}
graph TB
    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
    Main -->|contains| Priors["Prior Models"]
    Main -->|contains| Likelihoods["Likelihood Models"]
    Main -->|contains| Observations["Observation Models"]
    Main -->|contains| Guides["Inference Guides"]
    Main -->|contains| Registry["Registry System"]
    Main -->|contains| Factory["Factory System"]

    Dynamics -->|implements| DynamicsInterface["DynamicsModel Protocol"]
    Priors -->|implements| PriorsInterface["PriorModel Protocol"]
    Likelihoods -->|implements| LikelihoodsInterface["LikelihoodModel Protocol"]
    Observations -->|implements| ObservationsInterface["ObservationModel Protocol"]
    Guides -->|implements| GuidesInterface["InferenceGuide Protocol"]

    Factory -->|uses| Registry
    Factory -->|creates| Main

    class Main main;
    class Dynamics,Priors,Likelihoods,Observations,Guides component;
    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
    class Registry,Factory system;
```

### Core Components {#sec-core-components}

@@

#### Dynamics Models {#sec-dynamics-models}

Dynamics models implement the mathematical models of RNA dynamics, defining how unspliced and spliced RNA counts evolve over time.

- **StandardDynamicsModel**: The standard implementation of RNA dynamics
- **LegacyDynamicsModel**: Compatibility implementation matching the legacy model

```{mermaid}
classDiagram
    class DynamicsModel {
        <<Protocol>>
        +forward(context) Dict
        +steady_state(alpha, beta, gamma) Tuple
    }

    class StandardDynamicsModel {
        +name: str
        +description: str
        +shared_time: bool
        +t_scale_on: bool
        +forward(context) Dict
        +steady_state(alpha, beta, gamma) Tuple
    }

    class LegacyDynamicsModel {
        +name: str
        +description: str
        +shared_time: bool
        +t_scale_on: bool
        +forward(context) Dict
        +steady_state(alpha, beta, gamma) Tuple
    }

    DynamicsModel <|.. StandardDynamicsModel : implements
    DynamicsModel <|.. LegacyDynamicsModel : implements
```

#### Prior Models {#sec-prior-models}

Prior models define the prior distributions for model parameters, using Pyro's probabilistic programming framework.

- **LogNormalPriorModel**: Uses log-normal distributions for parameters

#### Likelihood Models {#sec-likelihood-models}

Likelihood models define the observation models, relating expected RNA counts to observed counts.

- **PoissonLikelihoodModel**: Uses Poisson distributions for RNA counts
- **LegacyLikelihoodModel**: Compatibility implementation matching the legacy model

#### Observation Models {#sec-observation-models}

Observation models handle the interface between data and model, preprocessing and transforming data.

- **StandardObservationModel**: The standard implementation for observations

#### Guide Factories {#sec-guide-factories}

Guide factories create guide functions for variational inference.

- **AutoGuideFactory**: Creates guides for the modular implementation
- **LegacyAutoGuideFactory**: Compatibility implementation matching the legacy guide

### Registry System {#sec-registry-system}

The registry system enables component registration and retrieval, implementing the registry pattern.

```{mermaid}
classDiagram
    class Registry~T~ {
        <<Generic>>
        +Dict _registry
        +register(name) Callable
        +get(name) T
        +create(name, **kwargs) T
        +list_available() List
        +clear()
    }

    class DynamicsModelRegistry {
        +Dict _registry
    }

    class PriorModelRegistry {
        +Dict _registry
    }

    class LikelihoodModelRegistry {
        +Dict _registry
    }

    class ObservationModelRegistry {
        +Dict _registry
    }

    class InferenceGuideRegistry {
        +Dict _registry
    }

    Registry <|-- DynamicsModelRegistry : extends
    Registry <|-- PriorModelRegistry : extends
    Registry <|-- LikelihoodModelRegistry : extends
    Registry <|-- ObservationModelRegistry : extends
    Registry <|-- InferenceGuideRegistry : extends
```

### Factory System {#sec-factory-system}

The factory system creates components and models from configurations, implementing the factory pattern.

```{mermaid}
classDiagram
    class ComponentFactory {
        +create_component(config, component_type) Component
        +create_component_from_dict(config_dict, component_type) Component
        +create_dynamics_model(config) DynamicsModel
        +create_prior_model(config) PriorModel
        +create_likelihood_model(config) LikelihoodModel
        +create_observation_model(config) ObservationModel
        +create_inference_guide(config) InferenceGuide
    }

    class ModelConfig {
        +ComponentConfig dynamics_model
        +ComponentConfig prior_model
        +ComponentConfig likelihood_model
        +ComponentConfig observation_model
        +ComponentConfig inference_guide
        +Dict metadata
    }

    class ComponentConfig {
        +str name
        +Dict params
    }

    ComponentFactory --> ModelConfig : uses
    ComponentFactory --> ComponentConfig : uses
```

## Module Organization {#sec-module-organization}

The modular implementation follows this organization:

```
src/pyrovelocity/models/modular/
├── __init__.py
├── comparison.py
├── components/
│   ├── __init__.py
│   ├── dynamics.py
│   ├── guides.py
│   ├── likelihoods.py
│   ├── observations.py
│   └── priors.py
├── config.py
├── constants.py
├── data/
│   ├── __init__.py
│   └── anndata.py
├── factory.py
├── inference/
│   ├── __init__.py
│   ├── config.py
│   ├── mcmc.py
│   ├── posterior.py
│   ├── svi.py
│   └── unified.py
├── interfaces.py
├── model.py
├── registry.py
├── selection.py
└── utils/
    ├── __init__.py
    ├── context_utils.py
    ├── core_utils.py
    └── pyro_utils.py
```

# Behavioral View {#sec-behavioral-view}

## Key Workflows {#sec-key-workflows}

### Model Creation Workflow {#sec-model-creation-workflow}

```{mermaid}
sequenceDiagram
    participant User
    participant Factory as ComponentFactory
    participant Registry as Registry System
    participant Components as Component Implementations
    participant Model as PyroVelocityModel

    User->>Factory: create_model_from_config(config)
    activate Factory
    Factory->>Factory: Extract component configs

    Factory->>Registry: create(dynamics_model.name, **dynamics_model.params)
    activate Registry
    Registry->>Components: Instantiate DynamicsModel
    Components-->>Registry: Return instance
    Registry-->>Factory: Return DynamicsModel
    deactivate Registry

    Factory->>Registry: create(prior_model.name, **prior_model.params)
    activate Registry
    Registry->>Components: Instantiate PriorModel
    Components-->>Registry: Return instance
    Registry-->>Factory: Return PriorModel
    deactivate Registry

    Factory->>Registry: create(likelihood_model.name, **likelihood_model.params)
    activate Registry
    Registry->>Components: Instantiate LikelihoodModel
    Components-->>Registry: Return instance
    Registry-->>Factory: Return LikelihoodModel
    deactivate Registry

    Factory->>Registry: create(observation_model.name, **observation_model.params)
    activate Registry
    Registry->>Components: Instantiate ObservationModel
    Components-->>Registry: Return instance
    Registry-->>Factory: Return ObservationModel
    deactivate Registry

    Factory->>Registry: create(inference_guide.name, **inference_guide.params)
    activate Registry
    Registry->>Components: Instantiate InferenceGuide
    Components-->>Registry: Return instance
    Registry-->>Factory: Return InferenceGuide
    deactivate Registry

    Factory->>Model: Create PyroVelocityModel with components
    Model-->>Factory: Return model instance
    Factory-->>User: Return model
    deactivate Factory
```

### Model Execution Workflow {#sec-model-execution-workflow}

```{mermaid}
sequenceDiagram
    participant User
    participant Model as PyroVelocityModel
    participant Observation as ObservationModel
    participant Prior as PriorModel
    participant Dynamics as DynamicsModel
    participant Likelihood as LikelihoodModel

    User->>Model: forward(u_obs, s_obs)
    activate Model
    Model->>Model: Initialize context

    Model->>Observation: forward(context)
    activate Observation
    Observation->>Observation: Preprocess data
    Observation-->>Model: Updated context
    deactivate Observation

    Model->>Prior: forward(context)
    activate Prior
    Prior->>Prior: Sample parameters
    Prior-->>Model: Updated context with parameters
    deactivate Prior

    Model->>Dynamics: forward(context)
    activate Dynamics
    Dynamics->>Dynamics: Compute expected counts
    Dynamics-->>Model: Updated context with expected counts
    deactivate Dynamics

    Model->>Likelihood: forward(context)
    activate Likelihood
    Likelihood->>Likelihood: Define likelihoods and observe data
    Likelihood-->>Model: Updated context with likelihoods
    deactivate Likelihood

    Model-->>User: Return results
    deactivate Model
```

## Component Lifecycle {#sec-component-lifecycle}

```{mermaid}
stateDiagram-v2
    [*] --> Registration: Component class defined
    Registration --> Configuration: User creates config
    Configuration --> Instantiation: Factory creates instance
    Instantiation --> Composition: Added to PyroVelocityModel
    Composition --> Execution: Model executed
    Execution --> [*]
```

# Architectural Decisions {#sec-architectural-decisions}

## ADR-1: Protocol-First Architecture {#sec-adr-1}

### Status

Accepted

### Context

The initial implementation used a mix of abstract base classes and Protocol interfaces, creating architectural inconsistency and forcing inheritance hierarchies.

### Decision

Adopt a Protocol-First architecture that completely removes base classes and standardizes exclusively on Protocol interfaces, with utility functions replacing base class functionality.

### Consequences

- Complete separation between interface and implementation
- Maximum flexibility in component implementation
- Full alignment with functional programming principles
- Zero coupling between components
- Greatly simplified testing and extension
- Temporary code duplication during transition, gradually reduced as utility functions are extracted

## ADR-2: Context-Based Component Communication {#sec-adr-2}

### Status

Accepted

### Context

Components need to share data and state during model execution.

### Decision

Use a shared context dictionary that is passed between components during model execution, with each component updating the context with its outputs.

### Consequences

- Flexible data sharing without tight coupling
- Clear data flow through the system
- Easy to extend with new data
- Potential for inconsistent context structure if not properly documented

# Cross-cutting Concerns {#sec-cross-cutting-concerns}

## Performance Considerations {#sec-performance-considerations}

PyroVelocity addresses performance through the following strategies:

- **Vectorized Operations**: Using PyTorch's vectorized operations for efficient computation
- **Batched Processing**: Supporting batch processing for large datasets
- **GPU Acceleration**: Leveraging PyTorch's GPU support for accelerated computation
- **Analytical Solutions**: Using analytical solutions where possible to avoid numerical integration

## Error Handling {#sec-error-handling}

The library follows these error handling principles:

- **Early Validation**: Validating inputs early to catch errors before computation
- **Informative Error Messages**: Providing clear error messages with context
- **Graceful Degradation**: Falling back to simpler models when advanced features fail

## Extensibility {#sec-extensibility}

The library is designed for extensibility through:

- **Protocol Interfaces**: Defining clear contracts for component implementations
- **Registry System**: Enabling dynamic registration of new components
- **Factory System**: Creating components from configurations
- **Composition**: Building complex models from simple components

# Technical Debt and Evolution {#sec-technical-debt}

## Known Limitations {#sec-known-limitations}

The current architecture has the following known limitations:

1. **Code Duplication**: Some duplication exists across component implementations
2. **Limited Documentation**: Documentation of component interfaces and behaviors is incomplete
3. **Validation Coverage**: Not all components have comprehensive validation against the legacy implementation

## Planned Improvements {#sec-planned-improvements}

The following architectural improvements are planned:

| Improvement | Priority | Timeline |
|-------------|----------|----------|
| Extract utility functions from duplicated code | High | Short-term |
| Improve documentation of component interfaces | High | Short-term |
| Expand validation coverage | Medium | Medium-term |
| Implement additional dynamics models | Medium | Medium-term |
| Enhance integration with AnnData | Medium | Medium-term |

# Implementation References {#sec-implementation-references}

## Key Source Files {#sec-key-source-files}

| Component | Primary Modules | Location |
|-----------|----------------|----------|
| Core Model | model.py | `src/pyrovelocity/models/modular/model.py` |
| Interfaces | interfaces.py | `src/pyrovelocity/models/modular/interfaces.py` |
| Dynamics Models | dynamics.py | `src/pyrovelocity/models/modular/components/dynamics.py` |
| Prior Models | priors.py | `src/pyrovelocity/models/modular/components/priors.py` |
| Likelihood Models | likelihoods.py | `src/pyrovelocity/models/modular/components/likelihoods.py` |
| Observation Models | observations.py | `src/pyrovelocity/models/modular/components/observations.py` |
| Guide Factories | guides.py | `src/pyrovelocity/models/modular/components/guides.py` |
| Registry System | registry.py | `src/pyrovelocity/models/modular/registry.py` |
| Factory System | factory.py | `src/pyrovelocity/models/modular/factory.py` |

## External Dependencies {#sec-external-dependencies}

| Dependency | Purpose | Impact |
|------------|---------|--------|
| PyTorch | Deep learning framework | Core computation engine |
| Pyro | Probabilistic programming | Core inference engine |
| AnnData | Data structure | Core data representation |
| Scanpy | Single-cell analysis | Data preprocessing and visualization |
| scVelo | RNA velocity analysis | Velocity computation and visualization |
| beartype | Runtime type checking | Type safety |
| jaxtyping | Type annotations for arrays | Shape information in type annotations |
| hydra-zen | Configuration management | Type-safe configuration |
